Classical mechanics as the high-entropy limit \\ of quantum mechanics

### 2025-12-09 Articles/2024-ClassicalLimit/classicalLimitArticle.tex
Clean copy for Physica Scripta, with corresponding author
Added hyphens, ~ after i.e., moved Wehrl entropy citation one sentence earlier, fixed refs (added preprint for one not yet published, removed months)

\title{Classical mechanics as the high-entropy limit \\ of quantum mechanics}
\author[1]{Gabriele Carcassi}
\author[2]{Manuele Landini}
\author[1]{Christine A. Aidala}

@@

For thermodynamic systems in equilibrium, entropy can be calculated directly based on occupation of the quantum states of the system. For high temperatures, the occupations are all small and classical statistical mechanics applies. Quantum effects related to indistinguishability are found whenever occupations approach unity. The entropy per particle drops in the same limit, again in line with our result. Note, however, that while infinite entropy corresponds to infinite temperature, the reverse is not necessarily true. In systems with finitely many possible states we can find the stationary point of $S$ as a function of $E$ for which $1/T=\partial S/\partial E$ is zero. This stationary point represents the maximum entropy reachable by the system, and therefore there is no high entropy limit. This is exactly the case where temperature can have negative values. Given that the entropy is a concave function of the energy, systems that admit a classical description are exactly those for which no such stationary point exists and temperature cannot be negative or infinite.%\footnote{{\color{blue}We can define temperature, based on the equation $1/T=\partial S/\partial E$. In systems with a finite number of degrees of freedom, we can find the stationary point of $S$ as a function of $E$. This is because both the minimum and maximum energy states will have $S=0$ or minimal. The maximum entropy state should correspond to infinite temperature. As a consequence, any microstate becomes equiprobable and the entropy takes the max value $S=k_B\log n_{dof}$} I reworked this point to again show how finitely many possibility does not give us a classical limit. }

@@

Having given qualitative reasons for the limit, and having seen how, in specific cases, standard arguments can be reinterpreted consistently with our view, we now develop the limit as a direct consequence of entropy increase. Since the maximum entropy attainable by a quantum system is the logarithm of the dimension of the corresponding Hilbert space~\cite{nielsen2010quantum}, the high-entropy limit exists only for infinite-dimensional spaces. This explains why spin, which lives in a finite-dimensional space, is an intrinsically quantum property. Therefore we are going to concentrate on the case of a single degree of freedom identified by position and momentum.\footnote{A similar limit should exist for an infinite-dimensional directional degree of freedom (i.e. for angular momentum). Though technically not a limit but more of an approximation, for finite-dimensional spaces one could look at mixed states that are close to the maximally mixed states (i.e. $\rho = I/n + \epsilon \sigma$).} The case of multiple degrees of freedom can be recovered by increasing the entropy of all DOFs at the same time, that is, by increasing the entropy of each DOF independently, without introducing correlations.

The overall argument can be broken down into the following steps:
\begin{enumerate}
    \item Characterize entropy-increasing maps in classical mechanics for a single DOF, noting that they can all be understood as Hamiltonian evolution followed by a pure stretching map, one that stretches position and momentum by the same factor $\sqrt{\lambda}$. Under this map, the expectation of a polynomial of position and momentum of degree $n$ is also multiplied by a factor of $\sqrt{\lambda^n}$.
    \item Show that a pure stretching map in quantum mechanics cannot be defined over symmetrized operator averages or normal ordering.
    \item Show that it can be defined over anti-normal ordering. Show that if we rescale the zero of entropy, the commutators become $[X, P]= \frac{\imath \hbar}{\lambda}$, which makes the limit $\lambda \to \infty$ mathematically equivalent to $\hbar \to 0$
    \item Show how, in the limit, the Wigner $W$ and Husimi $Q$ distributions become closer and closer.
\end{enumerate}

@@

For a pure stretching map, we can provide a more specific characterization. First of all, given that the map will act linearly on the space of probability distributions, once we characterize its action on distributions with compact support, we fully characterize the map. Furthermore, a distribution with compact support is always fully characterized by all its central moments, that is, the expectations for all polynomials of position and momentum (i.e. the moment problem is always solvable). Therefore, once we know how the central moments transform through the map, we know how all distributions with compact support transform and therefore we fully characterize the map. This means that $T$ is a pure stretching map if and only if
\begin{equation}
	\langle T(x^np^m) \rangle = (\sqrt{\lambda})^{(n+m)} \langle x^n p^m \rangle.
\end{equation}
Given the linearity of the expectation, this relationship will remain true for all distributions that have finite moments. In particular, this will be true for all Schwartz functions. Therefore, we can use the above as the defining property of the pure stretching map.

@@

It is interesting to consider what happens to the negative regions of $W$ under the stretching map. We know that $W$ can have negative regions, but their size is limited by the uncertainty principle. In fact, convolving $W$ with a 2D Gaussian with unitary spread, as in the definition of $Q$, returns a function that is never negative. In the limit $\lambda\gg1$, the formula for $W_\lambda$ reduces to
\begin{equation}
     W_\lambda(\beta)\rightarrow_{\lambda\gg1}\frac{2}{\pi\lambda^2}\int W_1\left(\frac{\alpha}{\sqrt{\lambda}}\right)e^{-\frac{2}{\lambda}\left|\alpha-\beta\right|^2}d^2\alpha=Q_\lambda(\beta).
\end{equation}
Therefore, while negative regions can be in principle found at any finite $\lambda$, $W$ tends to a positive function in the limit. As usual for the $W$ distribution, the phase space size of negative regions is limited to $\hbar$ by the uncertainty principle. The weight of the function in such regions is limited between $\pm 2/\hbar$ for pure states. The effect of the stretching map is to reduce this bound to $\pm 2/(\lambda\hbar)$ for large values of $\lambda$. A clear interpretation can be made by working directly on the Fourier transform of $W$; see Eq.~(\ref{Fourier}). The function $F(W_\lambda)$ is a scaled version of $F(W_1)$ with a Gaussian filter applied to it. Quantum information in $W_1$ is carried by spectral weights with k-vectors larger than 1. The bandwidth of the Gaussian filter is given by $\lambda/(\lambda-1)$. This cutoff approaches 1 in the limit, filtering away the interference terms.

@@

It should be clear that this investigation does not exhaust all possible insights and links to the now vast literature on open quantum systems, quantum statistical mechanics, quantum information, quantum sensing and so on. For example, entropy and entropy increase play a role in several approaches: coarse-grained measurement (e.g.~Refs.~\cite{Kofler2007,bibak2025classicallimitquantummechanics}), decoherence (e.g.~Refs.~\cite{Zurek_RevModPhys.75.715,DecoherenceBook2003,SCHLOSSHAUER20191}), stochastic processes (e.g.~Refs.~\cite{carmichael2013statistical,QuantumStochasticReview,QuantumFokkerPlanck2022}) and information-theoretic and resource-theoretic approaches (e.g.~Refs.~\cite{Goold_2016,QuantumResourceReview2019,Shiraishi2025}) to name a few. On a more technical level, since both the Husimi Q and the Wigner function converge to the classical distribution, the Wehrl entropy, which is the Shannon entropy calculated on the Husimi Q, will converge to the Gibbs entropy. One may ask, then, whether a high Wehrl entropy~\cite{Wehrl1979} is already an indicator of classicality, and, if so, it would give another insight as to why the anti-normal ordering is important. Lastly, the role of entropic ideas is well established in historic literature. It would not be possible here to comment on all of them, and the more we mention, the more the omitted mentions are felt. Therefore we only explore the most important elements to establish the limit mathematically and under the smallest conceptual footprint, also as to not distract from the main point of the paper.

@@

The limit also gives us additional insights. As we saw, classical states for which $\sigma_x \sigma_p < \frac{\hbar}{e}$ are states with negative entropy, which imply a breakdown of thermodynamics. It is no wonder, then, electrons cannot fall on the nucleus or that classical mechanics gives us the wrong spectra for black-body radiation. Classical mechanics fails at low entropy. The use of the correspondence principle as guidance for the development of quantum mechanics, then, can be understood as requiring that the new quantized theory reproduces classical mechanics at high entropy.

>>>>

\title{Classical mechanics as the high-entropy limit \\ of quantum mechanics}
\author[1]{Gabriele Carcassi}
\author[2]{Manuele Landini}
\author[1]{Christine A. Aidala\thanks{Corresponding author.~\href{mailto:caidala@umich.edu}{caidala@umich.edu}}}

@@

For thermodynamic systems in equilibrium, entropy can be calculated directly based on occupation of the quantum states of the system. For high temperatures, the occupations are all small and classical statistical mechanics applies. Quantum effects related to indistinguishability are found whenever occupations approach unity. The entropy per particle drops in the same limit, again in line with our result. Note, however, that while infinite entropy corresponds to infinite temperature, the reverse is not necessarily true. In systems with finitely many possible states we can find the stationary point of $S$ as a function of $E$ for which $1/T=\partial S/\partial E$ is zero. This stationary point represents the maximum entropy reachable by the system, and therefore there is no high-entropy limit. This is exactly the case where temperature can have negative values. Given that the entropy is a concave function of the energy, systems that admit a classical description are exactly those for which no such stationary point exists and temperature cannot be negative or infinite.%\footnote{{\color{blue}We can define temperature, based on the equation $1/T=\partial S/\partial E$. In systems with a finite number of degrees of freedom, we can find the stationary point of $S$ as a function of $E$. This is because both the minimum and maximum energy states will have $S=0$ or minimal. The maximum entropy state should correspond to infinite temperature. As a consequence, any microstate becomes equiprobable and the entropy takes the max value $S=k_B\log n_{dof}$} I reworked this point to again show how finitely many possibility does not give us a classical limit. }

@@

Having given qualitative reasons for the limit, and having seen how, in specific cases, standard arguments can be reinterpreted consistently with our view, we now develop the limit as a direct consequence of entropy increase. Since the maximum entropy attainable by a quantum system is the logarithm of the dimension of the corresponding Hilbert space~\cite{nielsen2010quantum}, the high-entropy limit exists only for infinite-dimensional spaces. This explains why spin, which lives in a finite-dimensional space, is an intrinsically quantum property. Therefore we are going to concentrate on the case of a single degree of freedom identified by position and momentum.\footnote{A similar limit should exist for an infinite-dimensional directional degree of freedom (i.e.~for angular momentum). Though technically not a limit but more of an approximation, for finite-dimensional spaces one could look at mixed states that are close to the maximally mixed states (i.e.~$\rho = I/n + \epsilon \sigma$).} The case of multiple degrees of freedom can be recovered by increasing the entropy of all DOFs at the same time, that is, by increasing the entropy of each DOF independently, without introducing correlations.

The overall argument can be broken down into the following steps:
\begin{enumerate}
    \item Characterize entropy-increasing maps in classical mechanics for a single DOF, noting that they can all be understood as Hamiltonian evolution followed by a pure stretching map, one that stretches position and momentum by the same factor $\sqrt{\lambda}$. Under this map, the expectation of a polynomial of position and momentum of degree $n$ is also multiplied by a factor of $\sqrt{\lambda^n}$.
    \item Show that a pure stretching map in quantum mechanics cannot be defined over symmetrized operator averages or normal ordering.
    \item Show that it can be defined over anti-normal ordering. Show that if we rescale the zero of entropy, the commutators become $[X, P]= \frac{\imath \hbar}{\lambda}$, which makes the limit $\lambda \to \infty$ mathematically equivalent to $\hbar \to 0$.
    \item Show how, in the limit, the Wigner $W$ and Husimi $Q$ distributions become closer and closer.
\end{enumerate}

@@

For a pure stretching map, we can provide a more specific characterization. First of all, given that the map will act linearly on the space of probability distributions, once we characterize its action on distributions with compact support, we fully characterize the map. Furthermore, a distribution with compact support is always fully characterized by all its central moments, that is, the expectations for all polynomials of position and momentum (i.e.~the moment problem is always solvable). Therefore, once we know how the central moments transform through the map, we know how all distributions with compact support transform and therefore we fully characterize the map. This means that $T$ is a pure stretching map if and only if
\begin{equation}
	\langle T(x^np^m) \rangle = (\sqrt{\lambda})^{(n+m)} \langle x^n p^m \rangle.
\end{equation}
Given the linearity of the expectation, this relationship will remain true for all distributions that have finite moments. In particular, this will be true for all Schwartz functions. Therefore, we can use the above as the defining property of the pure stretching map.

@@

It is interesting to consider what happens to the negative regions of $W$ under the stretching map. We know that $W$ can have negative regions, but their size is limited by the uncertainty principle. In fact, convolving $W$ with a 2D Gaussian with unitary spread, as in the definition of $Q$, returns a function that is never negative. In the limit $\lambda\gg1$, the formula for $W_\lambda$ reduces to
\begin{equation}
     W_\lambda(\beta)\rightarrow_{\lambda\gg1}\frac{2}{\pi\lambda^2}\int W_1\left(\frac{\alpha}{\sqrt{\lambda}}\right)e^{-\frac{2}{\lambda}\left|\alpha-\beta\right|^2}d^2\alpha=Q_\lambda(\beta).
\end{equation}
Therefore, while negative regions can be in principle found at any finite $\lambda$, $W$ tends to a positive function in the limit. As usual for the $W$ distribution, the phase-space size of negative regions is limited to $\hbar$ by the uncertainty principle. The weight of the function in such regions is limited between $\pm 2/\hbar$ for pure states. The effect of the stretching map is to reduce this bound to $\pm 2/(\lambda\hbar)$ for large values of $\lambda$. A clear interpretation can be made by working directly on the Fourier transform of $W$; see Eq.~(\ref{Fourier}). The function $F(W_\lambda)$ is a scaled version of $F(W_1)$ with a Gaussian filter applied to it. Quantum information in $W_1$ is carried by spectral weights with k-vectors larger than 1. The bandwidth of the Gaussian filter is given by $\lambda/(\lambda-1)$. This cutoff approaches 1 in the limit, filtering away the interference terms.

@@

It should be clear that this investigation does not exhaust all possible insights and links to the now vast literature on open quantum systems, quantum statistical mechanics, quantum information, quantum sensing and so on. For example, entropy and entropy increase play a role in several approaches: coarse-grained measurement (e.g.~Refs.~\cite{Kofler2007,bibak2025classicallimitquantummechanics}), decoherence (e.g.~Refs.~\cite{Zurek_RevModPhys.75.715,DecoherenceBook2003,SCHLOSSHAUER20191}), stochastic processes (e.g.~Refs.~\cite{carmichael2013statistical,QuantumStochasticReview,QuantumFokkerPlanck2022}) and information-theoretic and resource-theoretic approaches (e.g.~Refs.~\cite{Goold_2016,QuantumResourceReview2019,Shiraishi2025}) to name a few. On a more technical level, since both the Husimi Q and the Wigner function converge to the classical distribution, the Wehrl entropy~\cite{Wehrl1979}, which is the Shannon entropy calculated on the Husimi Q, will converge to the Gibbs entropy. One may ask, then, whether a high Wehrl entropy is already an indicator of classicality, and, if so, it would give another insight as to why the anti-normal ordering is important. Lastly, the role of entropic ideas is well established in historic literature. It would not be possible here to comment on all of them, and the more we mention, the more the omitted mentions are felt. Therefore we only explore the most important elements to establish the limit mathematically and under the smallest conceptual footprint, also as to not distract from the main point of the paper.

@@

The limit also gives us additional insights. As we saw, classical states for which $\sigma_x \sigma_p < \frac{\hbar}{e}$ are states with negative entropy, which imply a breakdown of thermodynamics. It is no wonder, then, electrons cannot fall on the nucleus or that classical mechanics gives us the wrong spectra for black-body radiation. Classical mechanics fails at low entropy. The use of the correspondence principle as guidance for the development of quantum mechanics, then, can be understood as requiring that the new quantized theory reproduce classical mechanics at high entropy.
